Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-09-12 RNARedPrint.tex

\author[
    addressref={aff1,aff2,aff3},
    email={stefan.hammer@uni-leipzig.de}
]{\inits{SH} \fnm{Stefan} \snm{Hammer}}
%
\author[
    addressref={aff4},
    email={wei.wang@lix.polytechnique.fr}
]{\inits{WW} \fnm{Wei} \snm{Wang}}
%
\author[
    addressref={aff2,aff3},
    email={will@tbi.univie.ac.at}
]{\inits{SW} \fnm{Sebastian} \snm{Will}}
%
\author[
    addressref={aff4,aff5},
    corref={aff4},                       % id of corresponding address, if any
    email={yann.ponty@lix.polytechnique.fr}
]{\inits{YP} \fnm{Yann} \snm{Ponty}}

@@

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}.$
An \Def{RNA (secondary) structure $R$ of
  length $n$} is a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
%
% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
% $i\leq i'\leq j \leq j'.$
%
\Def{Valid base pair} must pair bases from
$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$.

RNA energy models like the Turner energy model~\citep{Turner2009} define a function that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy}. 
Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops). It is most instructive, and in this context already practically useful, to develop our approach for the simple \Def{base pair energy model}. This model defines energy contributions $\EbpSym(x,y)$ for pairs $(x,y)\in\Sigma^2$ that specify the energy function as
$
\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
$
for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.

As the structural targets of our design, we fix the set of RNA structures
$\R:=\{R_1, \dots, R_k\}$ for sequences of length $n$.
Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases. Moreover, we want to enforce complex properties, which we express in terms of \emph{features}. An example of such a feature is the \GCb\% (the corresponding property of a sequence is to have a specific \GCb\%). The additional properties that we are going to enforce are specific 'target' energies for the single structures $R_1, \dots, R_k$; for this purpose, each such energy is considered as a feature.

Formally, we define a \Def{feature $F$} as function on sequences, which is expressed by summing over an associated set of \Def{feature functions} on sequences, such that
$$
F(S) = \sum_{\text{feature function $f$ associcated with $F$}} f(S).
$$
We emphasize that our approach is designed to profit (in terms of computational complexity) from feature functions that each depend only on few sequence positions. For example, the $\GCb\%$ feature is expressed by feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise. Similarly the base pair energy for a target structure $R$ is expressed as feature with feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j),$ such that $F(S)=\EbpSym(S,R)$. In both cases, the feature functions depend on the nucleotides at only a small (constantly bounded) number of sequence positions (one and two positions, respectively).

@@

The workhorse of our approach is the efficient (i.e.~fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
For weights $\pi_0,\dots,\pi_k$  compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_0,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

To stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
\begin{equation}
\label{eq:feature-energy-transformation}
\prod_{0\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
\end{equation}
where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features due to their additive nature.

\paragraph{Preliminaries on the complexity}

@@

In more generality, we consider dependencies due to the set of feature functions $\F$.
Recall that each feature function, even if evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.

@@

We are going to express our algorithms in the formalism of constraint networks, here
specialized as RNA design network. This abstraction allows us to base our
algorithm on the cluster tree elimination (CTE) of~\citet{Dechter2013}.

>>>>

\author[
    addressref={aff1,aff2,aff3},
    email={stefan.hammer@uni-leipzig.de}
]{\inits{SH} \fnm{Stefan} \snm{Hammer}}
%
\author[
    addressref={aff4},
    email={wei.wang@lix.polytechnique.fr}
]{\inits{WW} \fnm{Wei} \snm{Wang}}
%
\author[
    addressref={aff2,aff3},
    email={will@tbi.univie.ac.at}
]{\inits{SW} \fnm{Sebastian} \snm{Will$^*$}}
%
\author[
    addressref={aff4,aff5},
    corref={aff4},                       % id of corresponding address, if any
    email={yann.ponty@lix.polytechnique.fr}
]{\inits{YP} \fnm{Yann} \snm{Ponty}}

@@

An \Def{RNA sequence $S$} is a word over the \Def{nucleotide
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}.$
An \Def{RNA (secondary) structure $R$ of
  length $n$} is a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$ and for all different $(i,j), (i',j')\in R$:
$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
%
% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
% $i\leq i'\leq j \leq j'.$
%
\Def{Valid base pair} must pair bases from
$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\};$
we call $S$ \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$.

RNA energy models like the Turner energy model~\citep{Turner2009} define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$. 
Typically, such \Def{energy functions} are additively composed from \Def{energy contributions} of structure elements (like single base pairs or loops). It is most instructive, and in this context already practically useful, to develop our approach for the simple \Def{base pair energy model}. This model defines energy contributions $\EbpSym(x,y)$ for pairs $(x,y)\in\Sigma^2$ that specify the energy function as
$
\EbpSym(S,R) = \sum_{(i,j)\in R} \EbpSym(S_i,S_j)
$
for each sequence $S\in\Sigma^n$ and structure $R$ for length $n$.

As the structural targets of our design, we fix the set $\R:=\{R_1, \dots, R_k\}$ of $k$ RNA structures for sequences of length $n$.
Recall that we aim to generate random sequences that satisfy a number of constraints like the compatibility of paired bases. Moreover, we want to enforce complex properties, which we express in terms of \emph{features}. An example of such a feature is the \GCb\% (the corresponding property of a sequence is to have a specific \GCb\%). The additional properties that we are going to enforce are specific 'target' energies for the single structures $R_1, \dots, R_k$; for this purpose, each such energy is considered as a feature.

Formally, we define a \Def{feature $F$} as function on sequences, which is expressed by summing over an associated set of \Def{feature functions} on sequences, such that
$$
F(S) = \sum_{\text{feature function $f$ associcated with $F$}} f(S).
$$
We emphasize that our approach is designed to profit (in terms of computational complexity) from feature functions that each depend only on few sequence positions. For example, the $\GCb\%$ feature is expressed by $n$ feature functions $f^{GC}_i$ where $f^{GC}_i(S)$ equals 1 if $S_i='G'$ or $S_i='C'$ and 0 otherwise. Similarly the base pair energy for a target structure $R$ is expressed as feature with $|R|$-many feature functions $f^R_{ij}(S)=\EbpSym(S_i,S_j)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$. In both cases, the feature functions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively).

@@

The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
Given are the features $F_0,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple feature functions.
For weights $\pi_0,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_0,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{0\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

To stress the analogy to, in the RNA context more familiar (e.g.~\citep{McCaskill1990}), Boltzmann distributions over the structure ensemble of a sequence $S$, we emphasize the equivalence
\begin{equation}
\label{eq:feature-energy-transformation}
\prod_{1\leq \ell \leq k} \pi_\ell^{-F_\ell(S)} = exp(-\beta E(S)),
\end{equation}
where $\beta$ denotes inverse temperature and $E(S)$ is defined as linear combination of energies $E(S)=\sum_{1\leq \ell \leq k} \ln(\pi_\ell)/\beta E(S,R_\ell)$,
where the energy $E(S,R_\ell)$ is treated as feature $F_\ell(S) = \EbpSym(S,R_\ell);$ as elaborated later, even Turner energies or pseudoknot energies can be expressed as features due to their additive nature.

\paragraph{Further preliminaries for complexity and algorithmic considerations}

@@

In more generality, we consider dependencies due to the set of feature functions $\F$.
Recall that each feature function, even if `formally' evaluated for entire sequences, depends only on a subset of all sequence positions, which we call the \Def{dependencies of $f$} and write as $\dep(f)$;
thus, $\dep(f)$ is the minimal subset of $\{1,\dots,n\}$ s.t.~$\forall S,S'\in\Sigma^n: f(S)=f(S') \text{ only if } (\forall i\in\dep(f): S_i=S_i')$.

@@

We are going to express our algorithms in the formalism of constraint networks, here
specialized as RNA design network. This abstraction allows us to base our
algorithm on cluster tree elimination (CTE)~\citet{Dechter2013}.
